---
title: 'Case Study: How Does a Bike-Share Navigate Speedy Success?'
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Jamrus Ruttanayukol"
date: "`r as.Date('2022-Mar-23', format='%Y-%b-%d')`"
output: html_document
---

This page is a record of my Google Analytic program's capstone project. There were 2 case studies: bike-sharing company, and wellness company. I picked the former one.

Originally I wanted to do an open case study, but after looking around, I found out that there's not enough data for my interest - where should I buy a house in Thailand. Our census data is too outdated (only up to 2010, 2014, and 2017...). With so many big things going on recently, I lost interest. Maybe after I learn more I'll get back to it.

Anyway, Let's jump into the project! 

#### Organization 

This report is organized by Data analysis steps: 

 * Ask
 * Prepare
 * Process
 * Analyze
 * Share
 * Act

## Scenario

I'm a junior data analyst in marketing team at a Chicago bike-sharing company Cyclistic. The company wants to increase its annual membership. There are many questions to answer and they assigned me to look into how annual members and casual riders use our service differently. 

For more information, please click [here](https://d3c33hcgiwev3.cloudfront.net/aacF81H_TsWnBfNR_x7FIg_36299b28fa0c4a5aba836111daad12f1_DAC8-Case-Study-1.pdf?Expires=1648166400&Signature=H1lHhBbxoVuUHoufxSchQjKIrqBJNqXgVqBWC-VjtGhT4ojXScH~dQnMa9Twg1~5gGpKktzpZT07n1Dbue3g2BMZMhs6aS8N0eJayXZaDir0FHNaZssgqa3CF8ypt1LyKQJBRQt1tP-qQ77mSPFcwxProxG2XVbwFH0FbTE4Kgk_&Key-Pair-Id=APKAJLTNE6QMUY6HBC5A)

## Ask

#### Question: How do annual members and casual riders use Cyclistic bikes differently?

The end goal is getting more memberships. To this end, I need to spot behavioral differences that casual riders have from members. Especially ones that'd encourage them into subscriptions.

Once I spot any difficulties the casual riders face and how annual members do their cycling, I can recommend actions to my manager, Moreno. The team can also use my insights to improve their other topics of study.

### Deliverable

As a good practice, I should have a clear business task before diving in any data:

* Identifying the possible hurdles barring casual riders to join membership.

## Prepare

For this case study, we'll use the past 12 months trip data from Motivate International Inc. available [here](https://divvy-tripdata.s3.amazonaws.com/index.html). And before anything, I should note to myself that we must be aware of data-privacy issues and, at any circumstances, must not use customer's personally identifiable information (i.e. credit card numbers, home address, purchase history, etc.). For a blessing, the data does not contain any personal information unless I actively look for it. 

As this is an internal data, the original files will be kept in a local dedicated folder as I work in this project. Then, I'll upload transformed data into team's drive for other uses.

#### Data Credibility

A simple guideline for data credibility: ROCCC - Reliable, Original, Comprehensive, Current, and Cite.
The data seems OK. Since this data only involve company input (and automated data), input errors should be small.

#### Data Integrity

I looked at the data and see how it's formatted. Since this is in-house data, integrity should be good. 
Nonetheless, I'll confirm further as I move to next step.

### Deliverables

The data seems to contain all information needed for this analysis. We can check bicycle type, user type, where and when they started ride and ended the ride.

```{r set up environment, include=FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
library(hms)
library(ggplot2)
```

Let's check a dataset, say, 2021 March.

```{r import data and check column names}
tripdata_2103 <- read_csv('202103-divvy-tripdata.csv')
colnames(tripdata_2103)
```
Now let's take a quick look at the data.

```{r check each member record count in March 2021}
head(tripdata_2103)
tripdata_2103 %>%
  group_by(member_casual) %>%
  tally()
tripdata_2103 %>%
  group_by(rideable_type) %>%
  tally()
```
There seems to be no problem and data should be able to help us get the answers.

```{r import all 12 months of data, warning=FALSE, include=FALSE}
rm(tripdata_2103)
trip_2103 <- read_csv('202103-divvy-tripdata.csv')
trip_2104 <- read_csv('202104-divvy-tripdata.csv')
trip_2105 <- read_csv('202105-divvy-tripdata.csv')
trip_2106 <- read_csv('202106-divvy-tripdata.csv')
trip_2107 <- read_csv('202107-divvy-tripdata.csv')
trip_2108 <- read_csv('202108-divvy-tripdata.csv')
trip_2109 <- read_csv('202109-divvy-tripdata.csv')
trip_2110 <- read_csv('202110-divvy-tripdata.csv')
trip_2111 <- read_csv('202111-divvy-tripdata.csv')
trip_2112 <- read_csv('202112-divvy-tripdata.csv')
trip_2201 <- read_csv('202201-divvy-tripdata.csv')
trip_2202 <- read_csv('202202-divvy-tripdata.csv')
```

## Process

Now, onto the process part.

#### Chosen Tools

For this task, I'm planning to use only R. I've learned enough of SQL from UCDavis SQL basics specialization. I know Python is also useful but as of now I prefer to learn R. So, I hope I'll gain understanding of R basics through this course!

#### Data Integrity Assurance

This is in-house data from the system. So, we can trust that the data integrity attributes should be good. Nonetheless, we need to check so that we can be sure of our analysis results.

Note: data integrity - the accuracy, completeness, consistency, and trustworthiness of data throughout its life cycle.

```{r combine all 12 monthsdata}
trip_2022 <- rbind(trip_2201, trip_2202)
trip_2021 <- rbind(trip_2103, trip_2104, trip_2105, trip_2106, trip_2107, trip_2108, trip_2109, trip_2110, trip_2111, trip_2112)
trips <- rbind(trip_2021, trip_2022)
rm(trip_2021, trip_2022)
rm(trip_2103, trip_2104, trip_2105, trip_2106, trip_2107, trip_2108, trip_2109, trip_2110, trip_2111, trip_2112, trip_2201, trip_2202)
```

First, I'll do a simple checks for NULL and NA.

```{r check if any NULL or NA are there}
max(is.null(trips))
max(is.na(trips))
```
Oh it seems there's some NA. Let's filter and see.

```{r}
head(which(is.na(trips), arr.ind=TRUE))
```
OK. There's something wrong with column 5: "start_station_name".
Let's take a closer look.

```{r}
trips %>%
  filter(is.na(start_station_name)) %>%
  group_by(start_station_name, start_station_id) %>%
  tally()
```
Now it reveals that "start_station_id" also have NA. Let's delete these rows.

```{r remove rows with NA and re-check}
trips_c <- trips[complete.cases(trips),]
trips_c %>%
  filter(is.na(start_station_name)) %>%
  group_by(start_station_name, start_station_id) %>%
  tally()
```

Next let's confirm that we have no more NULL or NA.

```{r confirm cleanliness}
max(is.null(trips_c))
max(is.na(trips_c))
```

Great.
Next, we should re-arrange the data a bit. Let's add ride time and day of week columns. 

First, I'll add a column for time difference in seconds.

```{r add ride time}
trip_length <- difftime(as.POSIXct(trips_c$ended_at), as.POSIXct(trips_c$started_at), unit='secs')
trip_c1 <- mutate(trips_c, trip_length)
```

Next, I'll add a column for day of week.

```{r add day of week 0-6 starting from Sunday}
day_of_week <- as.POSIXlt(trips_c$started_at)$wday
trip_c2 <- mutate(trip_c1, day_of_week)
rm(trip_length, day_of_week, trips_c, trip_c1)
```

Now, trip_c2 should be good for use in Analyze phase.

### Deliverable

 * Documentation of any cleaning or manipulation of data --> done!

## Analyze

This is where all the fun begins!
Let's check descriptive stats.

```{r some basic stats 1}
mean(trip_c2$trip_length)
max(trip_c2$trip_length)
```
Certainly someone occupied their bike over 900 hours!
Let's see how many rows are with over 100 hours and what can I see from it.

```{r data with over 100 hours trip time}
filter(trip_c2, trip_length > 360000)
```
407 rows out of 463k. 
Let's leave it as it is for now.

Next, let's see day of week modes.

```{r day of week modes}
trip_c2 %>%
  group_by(day_of_week) %>%
  tally()
```

Most rides start on Saturday, but overall data is pretty well distributed.
Now let's go deeper. Firstly, average trip length for each user type.
```{r average trip length for each user type}
trip_c2 %>%
  group_by(member_casual) %>%
  summarize(mean(trip_length))
```

That's surprising. Casual users do spend more time on their rides. Let's dig more onto each day of week.
```{r average trip length by day of week for casual user 1}
trip_c2 %>%
  filter(member_casual == 'casual') %>%
  group_by(day_of_week) %>%
  summarize(mean(trip_length),n(),median(trip_length),max(trip_length),min(trip_length))
```

Let's look at member users data as well.
```{r average trip length by day of week for member user 1}
trip_c2 %>%
  filter(member_casual == 'member') %>%
  group_by(day_of_week) %>%
  summarize(mean(trip_length),n(),median(trip_length),max(trip_length),min(trip_length))
```

Strange. Both user types seems to have negative or zero trip length values. Did I miss something in the data?
```{r negative length?}
filter(trip_c2, trip_length <= 0)
```

Not many. Let's take them out.
```{r remove rows with negative trip time}
trip_c3 <- subset(trip_c2,!(trip_length <= 0))
```

And run the stats again.
```{r average trip length by day of week for casual user 2}
trip_c3 %>%
  filter(member_casual == 'casual') %>%
  group_by(day_of_week) %>%
  summarize(mean(trip_length),n(),median(trip_length),max(trip_length),min(trip_length))
```

```{r average trip length by day of week for member user 2}
trip_c3 %>%
  filter(member_casual == 'member') %>%
  group_by(day_of_week) %>%
  summarize(mean(trip_length),n(),median(trip_length),max(trip_length),min(trip_length))
```

Doesn't seem that helps with anything. 
Anyway, we can see more that casual users median and max trip time are higher than member users. 

Apparently, many casual members ride a bike on a daily basis. 
Marketing team could look further into this to get them subscribed!

### Deliverable

A summary of my analysis
I found 3 trends/relationships:

 * Casual users rode more on weekend - Member users trip count pretty much stayed constant throughout the week
 * Casual users rode longer than member users
 * Some casual users occupied their bike for months - not such thing happen with member users

## Share

Now I was just reminded that I will be presenting the findings to the executive team.
Let's make some viz to support my 3 findings. But first, I need to format my day of week correctly!

```{r re-format day of week column}
trip_c3$day_of_week <- recode(as.character(trip_c3$day_of_week), 
       "0"="Sunday",
       "1"="Monday",
       "2"="Tuesday",
       "3"="Wednesday",
       "4"="Thursday",
       "5"="Friday",
       "6"="Saturday")
trip_c3$day_of_week <- ordered(trip_c3$day_of_week, levels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday'))
```


```{r plot trips start day of week}
ggplot(data=trip_c3) + geom_bar(mapping=aes(x=day_of_week)) +
  facet_wrap('member_casual') + 
  labs(title = 'trips started in each day of week', subtitle = 'by casual users VS member users', x = 'day of week', y = 'trips' ) + theme(axis.text.x = element_text(angle = 45))
```

That's one viz down.
Next up, the average trip time.
```{r plot mean for each user type, message=FALSE, warning=FALSE}
ggplot(data=trip_c3) + stat_summary(aes(x = day_of_week, y = trip_length), fun = "mean", geom = "bar") +
  facet_wrap('member_casual') + 
  labs(title = 'Average trip duration started in each day of week', subtitle = 'by casual users VS member users', x = 'day of week', y = 'average trip duration (seconds)' ) + theme(axis.text.x = element_text(angle = 45))
```

That's another one done! Let's make the viz for the last one.

```{r plot max trip length for each user type}
ggplot(data=trip_c3) + stat_summary(aes(x = day_of_week, y = trip_length/3600), fun = "max", geom = "bar") +
  facet_wrap('member_casual') + 
  labs(title = 'Maximum trip duration started in each day of week', subtitle = 'by casual users VS member users', x = 'day of week', y = 'Maximum trip duration (hours)' ) + theme(axis.text.x = element_text(angle = 45))
```

Now that's all the viz created for all 3 findings. In actual work I'd move over to PowerPoint to make presentation with these visualizations (and some stats descriptions).

## Act

My top three recommendations would be:

* Highlight each users spending and how could they save by subscribing.
* Create offer to recommend how users can ride a bike to do their daily tasks
* Follow up with users who took bikes for over a month
